An Empirical Analysis of the Internet Engineering Task Force with Computational Methods

### 2024-10-03 paper.tex
Add content from Tracing Linguistic Markers of Influence... (ACL 2023)

\documentclass[twocolumn,10pt]{article}
\usepackage[l2tabu,orthodox]{nag}
\usepackage[utf8x]{inputenc}
\usepackage[british]{babel}
\usepackage{microtype}
\usepackage{amsmath}
\usepackage[all]{onlyamsmath}
\usepackage{newtxtext}
\usepackage{newtxmath}
\usepackage[caption=false]{subfig}
\usepackage{booktabs}
\usepackage{upquote}
\usepackage{graphicx}
\usepackage{url}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{color}
\usepackage[cm]{fullpage}

@@

\newcommand{\todo}[1]{\textbf{\textcolor{red}{To do: #1}}}
\newcommand{\pb}[1]{\vspace{0.75ex}\noindent{\textbf{#1}}}

@@

\title{An Empirical Analysis of the Internet Engineering Task Force 
       with Computational Methods}
\author{Mladen Karan \and
        Prashant Khare \and
        Matthew Russell Barnes \and
        Waleed Iqbal \and
        Junaid Qadir \and
        Stephen McQuistin \and
        Richard G. Clegg \and
        Patrick Healey \and
        Matthew Purver \and
        Ignacio Castro \and
        Gareth Tyson \and
        Colin Perkins}

@@

\pb{Summary:}
Increased \textit{Mobility} for working group chairs suggests that they
gain advantage from their role with their activity leading to similar
communication activity in the future. Moreover, they experience an
increased effect of \textit{Philanthropy} and \textit{Community}. The
former suggests that if a working group chair is engaged in a higher amount
of discussion, then in the future the individuals who they were
communicating with will also be highly active.  The latter suggests that
when working group chairs are surrounded by people who are active in
discussion, they are likely to be more active themselves in the future too.
All of this has less of an effect for regular participants who show a lower
correlation in terms of all three aspects. Our interpretation of this is
that active working group chairs are ``facilitating'' discussion in the WG
mailing lists; both having their activity boosted by their neighbours and,
in turn, boosting the activities of their neighbours. This aligns well with
the hypothesis that the cooperative and voluntary design of the IETF lends
itself well to higher levels encouraging those lower to engage in
discussion.

@@

% Tracing Linguistic Markers
% - RQ1: How do linguistic traits differ between more and less influential
%   participants? 
% - RQ2: How do linguistic traits vary for participants at different levels
%   of the organisation hierarchy? 
% - RQ3: How does linguistic behaviour of participants change as they gain
%   influence?

@@

\subsection{What Makes a Successful Author?}
% ICWSM 2022 paper Section 4

@@

\subsection{What Makes a Successful Document?}

>>>>

\documentclass[twocolumn,10pt]{article}
\pdfoutput=1 % This must be in the first 5 lines to tell arXiv to use pdfLaTeX
\usepackage[l2tabu,orthodox]{nag}
\usepackage[utf8x]{inputenc}
\usepackage[british]{babel}
\usepackage{microtype}
\usepackage{amsmath}
\usepackage[all]{onlyamsmath}
\usepackage{newtxtext}
\usepackage{newtxmath}
\usepackage[caption=false]{subfig}
\usepackage{booktabs}
\usepackage{upquote}
\usepackage{graphicx}
\usepackage{url}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{color}
\usepackage[cm]{fullpage}
\usepackage{multirow}

@@

\newcommand{\todo}[1]{\textbf{\textcolor{red}{To do: #1}}}
\newcommand{\pb}[1]{\vspace{0.75ex}\noindent{\textbf{#1}}}
\newcommand{\liwc}[1]{\textsc{#1}}
\newcommand{\word}[1]{\textit{`#1'}}

@@

\title{An Empirical Analysis of the Internet Engineering Task Force 
       with Computational Methods}
\author{Mladen Karan \and
        Prashant Khare \and
        Ravi Shekhar \and
        Matthew Russell Barnes \and
        Waleed Iqbal \and
        Junaid Qadir \and
        Stephen McQuistin \and
        Richard G. Clegg \and
        Patrick Healey \and
        Matthew Purver \and
        Ignacio Castro \and
        Gareth Tyson \and
        Colin Perkins}

@@

%..................................................................................................
\pb{Summary:}
Increased \textit{Mobility} for working group chairs suggests that they
gain advantage from their role with their activity leading to similar
communication activity in the future. Moreover, they experience an
increased effect of \textit{Philanthropy} and \textit{Community}. The
former suggests that if a working group chair is engaged in a higher amount
of discussion, then in the future the individuals who they were
communicating with will also be highly active.  The latter suggests that
when working group chairs are surrounded by people who are active in
discussion, they are likely to be more active themselves in the future too.
All of this has less of an effect for regular participants who show a lower
correlation in terms of all three aspects. Our interpretation of this is
that active working group chairs are ``facilitating'' discussion in the WG
mailing lists; both having their activity boosted by their neighbours and,
in turn, boosting the activities of their neighbours. This aligns well with
the hypothesis that the cooperative and voluntary design of the IETF lends
itself well to higher levels encouraging those lower to engage in
discussion.

@@

Moving on from discussing the people involved in the IETF (\S\ref{sec:trends-demographics})
and their interactions (\S\ref{sec:org-dyn}), we now consider communications
within the IETF from the perspective of language to explore whether there
are differences in linguistic patterns and language use in more influential
participants, and to consider how the use of language reflects the
consensus-driven nature of the standards development process.

% - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
% The following is adapted from our ACL 2023 Tracing Linguistic Markers of
% Influence paper, section 2.

\pb{Linguistic Inquiry and Word Count} (LIWC) \cite{pennebaker2015development}
is a well-recognised psycholinguistic lexicon; it provides word counts for
85 different linguistic, psychological, personal concern, and informal
language marker categories. We generate a \emph{LIWC representation} of
the emails sent by a participant in the analysis that follows. This is
calculated by aggregating the word counts within each linguistic category
for each participant using the LIWC 2015 dictionary (academic license).
We filter out 104 ambiguous words that are present in LIWC but have
technology, security, and network context meaning in IETF, using manually
curated lists, for e.g., attack, argument, secure etc.,

We also normalise by the total number of emails sent by that participant. 
Such a normalisation is more appropriate here than normalising by total
number of words written, as many IETF emails include long technical
sections.  This generates a representation of a participant as their mean
usage of each LIWC category; while this is a relatively reduced,
low-dimensional representation of a person's language, it has the advantage
of being interpretable and psychologically well-motivated.

A limitation is that we used the standard LIWC-based analysis approach,
which is purely lexical and does not take into account the context in which
a word appears. Consequently, many words that have very specific senses in
the context of the IETF get miscounted as occurrences of LIWC categories.
This could be addressed in future by a more advanced method of mapping to
LIWC categories that would account for context.

%..................................................................................................
% The following is from our ACL 2023 Tracing Linguistic Markers of
% Influence paper RQ1: How do linguistic traits differ between more
% and less influential participants?

\pb{Language and Influence:} How do linguistic traits differ between more
and less influential participants? That is, are there differences in the
way influential participants, based on centrality in the email social
graph, make use of language compared to those who are less influential?

We study mailing list data from the period 2015-2019, comprising 300,806
emails from 5,363 unique participants. We used a 5-year subset of the data
due to the computation cost, while still giving a reasonable period to
observe the participation consistency in the IETF community, and chose the
last period prior to the distortion of the pandemic.  For each participant,
we calculate the centrality-based influence score and LIWC representation,
We fit a linear regression model using LIWC representations to predict
influence percentile and observe the magnitude and directions of
significant coefficients.

The following LIWC categories are highly correlated ($p < 0.05$) with
centrality-based influence: \liwc{we, informal, risk, adjective, anger,
they}, and \liwc{bio}.  Categories such as \liwc{netspeak, sexual, health,
death, body} are correlated with lower influence. This suggests that
influential people tend to indicate a collaborative and community-oriented
approach with first-person plural (\liwc{we}) and third-person plural
category (\liwc{they}) usage. This is consistent with
\cite{kacewicz2014pronoun} and \cite{guinote2017power}, who show that
influential people use more first-person plural. They also use more
organisational language, which is shown by the negative correlation of
informal slang language categories (\liwc{netspeak, sexual, body}). We see
some unexpected hidden trends due to word ambiguity (e.g., words like
`\textit{trust}' and `\textit{live}'), which are discussed below.

%..................................................................................................
% The following is from our ACL 2023 Tracing Linguistic Markers of
% Influence paper RQ2: How do linguistic traits vary for participants
% at different levels of the organisation hierarchy?

\pb{Language and Organisational Role:}
How do linguistic traits vary for participants at different levels of the
organisation hierarchy?

Centrality-based influence considers participants based on their email
network, but role-based influence, based on the position of a person in the
organisational hierarchy (regular participant, working group chair, area
director) is equally crucial as they are involved in organisational
decision making.\footnote{In the top 10\% \textit{mail-based} influential
participants, less than 30\% are WG chairs with significant
\textit{role-based} influence.} 
For the same time period, 2025-2019, we split the data into two categories:
(a) emails sent by working group chairs, and (b) emails sent by regular
participants who have never been a working group chair (the number of area
directors is small, so we exclude them from this analysis).  We calculate
the LIWC representations for each person, train a logistic regression model
to predict category, and observe the LIWC category coefficients.

From \ref{table:all_results}, we see that working group chairs are more
social and collaborative, as is shown by \liwc{we} and \liwc{social}
categories. This is in line with our findings for influence-based
centrality above and resuslts on leadership engagement from the literature
\cite{strzalkowski2012modeling, liu2022pronoun, kacewicz2014pronoun,
guinote2017power}.
Chairs also use more tentative statements (\liwc{tentat}) in discussions,
primarily focused on technical feedback and revisions, or suggesting
alternatives. For example:
\textit{``With the risk of disturbing with statements, but avoiding too
many questions:This \underline{seems} against the goal of reducing
headers.''} 
and
\textit{``Question is do we need to carry around an outer IP-in-IP header
for that \underline{or} not?''}.

\begin{table*}[]
  \small
  \begin{center}
  \begin{tabular}{l|l|l}
  \toprule
    \multirow{2}{*}{Language and influence} 
      & High influence
      & \liwc{bio, we, informal, they, negemo, anger, risk, adjective}  \\ 
    \cmidrule{2-3}
      & Low influence
      & \liwc{sexual, death, ingest, netspeak, health, female, body, affiliation, conj} \\ 
  \hline
    \multirow{2}{*}{Language and role} 
      & WG Chair influence
      & \liwc{tentat, ipron, social, see, feel, we} \\
    \cmidrule{2-3}
      & non-WG Chair
      & \liwc{cogproc, relativ, affiliation, i, reward} \\
  \hline
    \multirow{2}{*}{Changes in Language}
      & Top 10 percentile
      & \liwc{adverb, prep, anger, auxverb, male, cogproc, achiev, risk, focuspresent} \\
    \cmidrule{2-3}
      & Below 50$^{th}$ percentile
      & \liwc{function, ppron, shehe, ipron, number, certain, sexual, informal} \\
  \hline
  \end{tabular}
  \end{center}
  \caption{LIWC categories where $p<0.05$.}
  \label{table:all_results}
\end{table*}

%..................................................................................................
% The following is from our ACL 2023 Tracing Linguistic Markers of
% Influence paper RQ3: How does linguistic behaviour of participants
% change as they gain influence?

\pb{Changes in Use of Language with Changes in Role:}
Finally, we look at how linguistic behaviour of participants changes as
they gain influence in the IETF.
We consider at participants who went from low to high influence over time:
individuals who had a \emph{centrality-based} influence below the 50th
percentile when they joined the IETF, and reached the top 10th percentile
at some point.

For each participant, we generate two different representations based on
two periods---the year of joining, and year of reaching the top 10th
percentile for the first time---and assign these to two different classes.
We then train a logistic regression model to predict these classes, and
examine the coefficients of the LIWC categories

From Table \ref{table:all_results}, we observe that when participants
become influential, according to the centrality-based metric, they are
likely to be more descriptive and engaged in immediate state of issues
and situations as seen from the correlation of auxiliary verbs
(\liwc{auxverb}), adverb, risk, and present focus (\liwc{focuspresent}).
They are also more involved in cognitive processes (\liwc{cogproc}) as
compared to their previous self when they were new to IETF and had little
influence.

%..................................................................................................
\pb{Discussion:}
To better understand these LIWC categories and what kind of words play a
role in the behaviour of individual categories, we calculate the frequency
of words in each LIWC category as they appear in the emails. 
Next, we consider the top 30 most frequent words in each LIWC category and
perform regression analysis on centrality-based influence for participants,
but using only these 30 words as features to generate the participant
representation. We conducted this experiment separately for each LIWC
category that was significant in the first experiment.

From the word based analysis we make multiple observations. Firstly, words
like \word{we} imply a collective approach and is strongly correlated with
the higher influence. 
Similarly, the use of word \word{well} is standard, such as politely
resuming the conversation (e.g., `\textit{well, I agree}') or providing an
approval over something (e.g., `\textit{this works as well}'). These words
are well associated with the influential participants.  Otherwise,
influential participants are generally not observed to be informal and
other frequent words (other than \word{well}) within \liwc{informal}
category do not demonstrate a strong correlation with the growing
influence. Also, \word{well} is the most frequent word in the
\liwc{informal} category.

More influential people, according to both centrality- and role-based
metrics, are also observed to engage more with the community.
Conversations can often reflect situations where, as a part of review and
feedback process, more influential people highlight limitations in protocol
standards, stress on specifics, and compare with existing protocols or
previous versions.
Several words across different LIWC categories (\liwc{risk}, \liwc{negemo},
and \liwc{adj}) highlight such behaviour, e.g., \word{problems}, \word{before},
\word{particular}, \word{specific}, \word{different}, \word{most}, and \word{than}.

However, there are many words with dual sense, like \word{trust} which has
a very technology specific usage related to network security instead of
conversations involving trust issues between individuals or trust in any
given situation. Similarly, the word \word{live} is related with an
application or network being live, instead of its conventional meaning. We
also observed that some of the LIWC categories, such as \liwc{bio}, did not
have specific terms that could clearly establish its significance in favour
of influential participants (e.g., word \word{problems} and \word{trust}
reflecting the significance for the category \liwc{risk}), instead such
categories had several words with quite weak correlation with influential
participants. Such words collectively drifted the weight of the category
towards influential participants.

%..................................................................................................
\pb{Summary:}
Using two aspects of influence, based on centrality in the email social
graph and based on organisation role, we were able to unfold several traits
that differentiate influential participants from others. Many of our
findings seem corroborated by studies in organisational theory. 
We observed that influential people exhibit more collaborative and
community-oriented traits, and also stronger signs of engagement in
discussions. We also observed that as people go on to become influential
participants, they evolve in their communication and are seen to be more
engaging and descriptive in their linguistic style.

% - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -

@@

%--------------------------------------------------------------------------------------------------
\subsection{What Makes a Successful Author?}
% ICWSM 2022 paper Section 4

@@

%--------------------------------------------------------------------------------------------------
\subsection{What Makes a Successful Document?}
